Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-01-19 paper-ieee-trans.tex

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
which supports an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One of the well-known examples is the Worldwide 
Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered 
by the Particle Physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the Large 
Hadron Collider particle accelerator. A significant number of innovation projects and
technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
and ultimately by the Horizon 2020 programme\footnote{https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
develop distributed solutions that provide a seamless and transparent access to those
computing and data resources, available for the researchers.

@@

Analyzing the European Commission (EC)-funded software projects in the last decade
we have observed a continuous increase of the prominence and robustness of
testing and validation procedures. One of many reasons is related to the
evolution of the software engineering methodologies (e.g. creation of the Agile
Manifesto \cite{agile-manifesto}, rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communications technology (ICT)
field such as virtualization, enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization. All of which
complemented by the emergence of automation and event-response tools.

@@

The EC projects considered in this text are listed chronologically in Table
\ref{tab:eup}. In most cases, each project paved the way of the subsequent projects,
following a logic evolution in which the quality and reliability of software have
increasingly had a more significant role. Experiences gathered during the 
course of each project support this fact and, as it is shown in Table \ref{tab:feat},
latest projects are the most committed in terms of software reliability awareness.
This table summarizes the topics addressed in the following sections, highlighting 
the commonalities and specific features on software reliability.

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC-projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2004\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E-sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI-Inspire &
Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO-DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
had as main goal to develop the software to provide basic grid functionality
and associated management tools for a large scale testbed for demonstration projects in three
specific areas of science, such as Particle physics, Bioinformatics and Earth observation.
%was devoted to develop the first European infrastructure for Grid computing. 
The project brought together 21 academic and industry partners, from 15 different
countries. 
%There were 3 application areas for requirement gathering; Particle physics, Bioinformatics and Earth observation. 
The project delivered its own software distribution, named EDG (EU DataGrid), strongly 
based on Globus Middleware Services \cite{globus}. The project was organized in 12 workpages, from 
which 5 were devoted to software development coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

@@

The source code was available at a private, centralized version control system.
The code was passed through a manual certification procedure at the time of the release. 
This procedure tried to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process 
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
production. Starting with EGEE--II, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware: ETICS 
\cite{etics} solution.

@@

EGI--InSPIRE not only ensured the coordination of the operation of a secure, reliable 
European-wide production grid infrastructure federated from national grid initiatives 
that is integrated and interoperates with other grids worldwide but also foreseed the 
maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
the EGI Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
coordination of the external provision of software required by EGI to 
form the production infrastructure. For the later it maintained a production--ready 
Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution) and 
defined general and component specific quality criteria to be applied to software 
components made available through this repository.
The role of {\sl UMD} was to enforce the fulfilment of this set of quality
criteria \cite{egi-qc} in the software being delivered (i.e.,
{\sl EMI} and {\sl Globus}) \cite{mario}. {\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under a follow--up
project, the EGI--Engage (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
the project is succeeding in providing solutions to address the existing gaps in
cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

The INDIGO--DataCloud project is the most recent project in the list and, as such, it is 
the result of the lessons learned from previous experiences in terms of the evaluation and application
of innovative software engineering methodologies, the management of new technologies to put
them into practice, and the compilation and analysis of metrics to measure the success of the
solutions implemented. The expertise gathered is complemented by the fact of being framed in
collaborative projects with an extensive list of partners.

framed projects

All of this with the added complexity of being framed in major
collaborations with

procollecting and analyzing metrics relevant for the .. to empower the sw reliability bla bla bla

The INDIGO--Datacloud project reflects the progress made in software
quality and reliability aspects throughout the past scientific European
experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
highly influenced by the insights of current big worldwide collaborations of
software development. These collaborations have continuously inspired the use of new
technologies and procedures to increase the robustness and quality of the
software being delivered. In this scenario, the implementation of the DevOps culture 
is one of the major outcomes, as this paradigm has been progressively adopted by the
project.

The DevOps methods emphasize on exercising Quality Assurance techniques to
avoid infrastructure disruption whenever new developments are deployed into
production. The INDIGO--DataCloud project promoted the application of a 
Continuous Integration (CI) scenario to enforce the Quality Assurance requirements 
for any piece of software being produced within the project. Such environment 
requires an automation ready--to--go infrastructure where the different services involved
(source code management platform, automation server) interact with each other
to trigger the quality check execution and return back the exit codes, together
with their outputs. To accomplish this, the project leveraged on tightly
integrated open source tools such as GitHub \cite{github}, as source code repository
and Jenkins \cite{jenkins}, as continuous integration system, as well as relying on
Docker container provisioning to speed up the source code validation. This CI
approach is guiding the project’s software development phase throughout the
first and second major releases.

As DevOps suggests, frequent releases positively affect the reliability of the
software. The software updates of INDIGO--DataCloud products taking place since
the second major release are passing through a Continuous Delivery (CD)
pipeline that adds the packaging of the software right after the successful execution of
the quality checks (as part of the CI). 
The steps defined within the CD pipeline differ whether the software is to be distributed via Docker
images (see Figure \ref{fig:fig_CD}) or via operating system’s packages, rpms or debs. In the latter case an extra
validation step is added, consisting in the product’s deployment using a
Configuration Management (CM) solution. The installation uses the packages
built in the previous step and uploaded to a testing
repository. In the case of Docker images, the CM is used to build the image
itself, installing and configuring the product before uploading it to the
DockerHub repository \cite{indigo-dockerhub}.

\subsection{Software Quality Procedures}

@@

@@

\subsection{DevOps adoption from user communities}

@@

\bibitem{condor}
D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the grid}, in Grid
Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70,
Eds. John Wiley \& Sons Inc., 2002.

@@

\bibitem{etics}
A. Di Meglio, M.-E.Bégin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
the international software engineering service for the grid}, in Journal of
Physics: Conference Series, Vol. 119, N. 4, IOP Publishing Ltd, 2008.

@@

% if you will not have a photo at all:
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/pablo-orviz}}]{Pablo Orviz Fernandez}
is a computing researcher at CSIC (Consejo Superior de Investigaciones
Cientificas), holding an MSc in Scientific Computing from the University of
Cantabria. During the past years he has been devoted to software quality
assurance tasks within different European advanced computing research projects
such as INDIGO-DataCloud and EGI-Engage. Currently involved in EOSC-hub project
and participating in the software quality work packages of DEEP-HybridDatacloud
and eXtreme-DataCloud.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/mario-david}}]{Mario David}
is a research associate at LIP. He holds a PhD in Experimental Particle Physics from the University of Lisbon.
Member of DEEP-HybridDatacloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC-hub project and
of the Portuguese National Computing Distributed Infrastructure.
He held a research associate position at Institut de Physique du Globe de Paris (IPGP/CNRS) as Scientific Software Developer for the VERCE project, in particular in the data intensive use cases for seismology.
He was been actively involved in the Validation, Quality Assurance and testing of middleware, in regional and global operations.
\end{IEEEbiography}

>>>>

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
which supports an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One of the well-known examples is the Worldwide 
Large Hadron Collider (LHC) Computing Grid (WLCG) global computing infrastructure, powered 
by the particle physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the Large 
Hadron Collider particle accelerator. A significant number of innovation projects and
technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
and ultimately by the Horizon 2020 programme\footnote{HORIZON 2020 - The EU Framework
Programme for Research and Innovation https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
develop distributed solutions that provide a seamless and transparent access to those
computing and data resources, available for the researchers.

@@

Analyzing the European Commission (EC)-funded software projects in the last decade
we have observed a continuous increase of the prominence and robustness of
testing and validation procedures. One of many reasons is related to the
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology (ICT)
field (such as virtualization, the enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization). All of which
complemented by the emergence of automation and event-response tools.

@@

The EC projects considered in this text are listed chronologically in Table
\ref{tab:eup}. In most cases, each project paved the way of the subsequent projects,
following a logic evolution in which the quality and reliability of software have
increasingly had a more significant role. Experiences gathered during the 
course of each project support this fact and, as it is shown in Table \ref{tab:feat},
latest projects are the most committed in terms of software reliability awareness.
Table \ref{tab:feat} summarizes also the topics addressed in the following sections, highlighting 
the commonalities and specific features on software reliability.

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC-projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E-sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI-Inspire &
Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO-DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
had as main goal to develop the software to provide basic Grid functionality
and associated management tools for a large scale testbed for demonstration projects in three
specific areas of science, such as particle physics, bioinformatics and earth observation.
%was devoted to develop the first European infrastructure for Grid computing. 
The project brought together 21 academic and industry partners, from 15 different
countries. 
The project delivered its own software distribution, named EDG (EU DataGrid), strongly 
based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
which 5 were devoted to software development coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

@@

The source code was available at a private, centralized version control system.
The code was passed through a manual certification procedure at the time of the release. 
This procedure tried to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process 
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
production. Starting with EGEE--II, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the ETICS \cite{etics} solution.

@@

EGI--InSPIRE not only ensured the coordination of the operation of a secure, reliable 
European-wide production Grid infrastructure federated from National Grid Initiatives (NGIs)
that is integrated and interoperates with other Grids worldwide but also foreseed the 
maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
the EGI Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
coordination of the external provision of software required by EGI to 
form the production infrastructure. For the later it maintained a production--ready 
Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution), and 
defined general and component specific quality criteria to be applied to software 
components made available through this repository.
The role of {\sl UMD} was to enforce the fulfilment of this set of quality
criteria \cite{egi-qc} in the software being delivered (i.e.,
{\sl EMI} and {\sl Globus}) \cite{mario}. {\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under a follow--up
project, the EGI--Engage (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

INDIGO--DataCloud is the most recent project included in this paper thus it
emerged from the lessons learned in the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years in terms
of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of metrics to measure the
success of the solutions implemented, has been complemented by the added 
complexity of being framed into large collaborative projects with an extensive
list of partners.

Three main pillars sustain the project's strategy to seek reliability in the 
software produced. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} guides the development of the software in
a per-change basis, validating the software at the very early stages. The 
software delivery process is designed as a bottom-up \textit{adoption of DevOps
principles}, passing through Continuous Integration (CI) and Continuous
Delivery (CD) approaches. Once the software is released, it follows a 
\textit{two-factor validation} in the project's pilot infrastructures and
resource provider's early-adoption.

%The INDIGO--Datacloud project reflects the progress made in software
%quality and reliability aspects throughout the past scientific European
%experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
%highly influenced by the insights of current big worldwide collaborations of
%software development. These collaborations have continuously inspired the use of new
%technologies and procedures to increase the robustness and quality of the
%software being delivered. In this scenario, the implementation of the DevOps culture 
%is one of the major outcomes, as this paradigm has been progressively adopted by the
%project.

\subsection{Software Quality Procedures}
\label{subsec:sqa}

@@

The DevOps methods emphasize on exercising SQA techniques to avoid infrastructure
disruption whenever new developments are deployed into production systems. The 
following DevOps approaches have been progressively adopted throughout the project's
lifetime, being one of its major outcomes.

\subsubsection{Continuous Integration}
The INDIGO--DataCloud project promoted the application of a CI scenario to enforce
the SQA requirements -- described in Section \ref{subsec:sqa} -- for any piece of
software being produced within the project. Such environment requires an automation
ready--to--go infrastructure where the different technologies involved -- source code
management platform, automation server and containerization -- interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is comprised of diverse quality checks, enforcing at least the successful
execution of code style and minimum unit testing coverage. To take over the 
implementation of such infrastructure, the project leveraged on tightly integrated open
source tools such as GitHub \cite{github}, as the online source code repository,
and Jenkins \cite{jenkins}, the event-response CI system. Docker container provisioning
is used to provide instant computing power needed for executing the pipeline jobs.

This CI approach is guiding the project’s software development phase throughout the
first and second major releases.

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the
software. The software updates of INDIGO--DataCloud products taking place since
the second major release are passing through a Continuous Delivery (CD)
pipeline that adds the packaging of the software right after the successful execution of
the quality checks (as part of the CI). 
The steps defined within the CD pipeline differ whether the software is to be distributed via Docker
images (see Figure \ref{fig:fig_CD}) or via operating system’s packages, rpms or debs. In the latter case an extra
validation step is added, consisting in the product’s deployment using a
Configuration Management (CM) solution. The installation uses the packages
built in the previous step and uploaded to a testing
repository. In the case of Docker images, the CM is used to build the image
itself, installing and configuring the product before uploading it to the
DockerHub repository \cite{indigo-dockerhub}.

@@

\subsubsection{DevOps adoption from user communities}

@@

\bibitem{condor}
D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the Grid}, in Grid
Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70,
Eds. John Wiley \& Sons Inc., 2002.

@@

\bibitem{etics}
A. Di Meglio, M.-E.Bégin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
the international software engineering service for the Grid}, in Journal of
Physics: Conference Series, Vol. 119, N. 4, IOP Publishing Ltd, 2008.

@@

% if you will not have a photo at all:
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/pablo-orviz}}]{Pablo Orviz Fernandez}
is a computing researcher at CSIC (Consejo Superior de Investigaciones
Cientificas), holding an MSc in Scientific Computing from the University of
Cantabria. During the past years he has been devoted to software quality
assurance tasks within different European advanced computing research projects
such as INDIGO-DataCloud and EGI-Engage. Currently involved in EOSC-hub project
and participating in the software quality work packages of DEEP-HybridDataCloud
and eXtreme-DataCloud.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/mario-david}}]{Mario David}
is a research associate at LIP. He holds a PhD in Experimental Particle Physics from the University of Lisbon.
Member of DEEP-HybridDataCloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC-hub project and
of the Portuguese National Computing Distributed Infrastructure.
He held a research associate position at Institut de Physique du Globe de Paris (IPGP/CNRS) as Scientific Software Developer for the VERCE project, in particular in the data intensive use cases for seismology.
He was been actively involved in the Validation, Quality Assurance and testing of middleware, in regional and global operations.
\end{IEEEbiography}
